---
title: "Initial Tables"
bibliography: reference.bib
editor_options: 
  chunk_output_type: inline
---

# By Play

```{r setup, message=FALSE}
#| code-fold: true
#| code-summary: Imports

library(nflverse)
library(tidyverse)
library(ggplot2)
library(grid)
library(reticulate)
library(kableExtra)
use_condaenv("r-env")
options(scipen = 999)

# raw_nfl <- load_4th_pbp(2016:2022)
```

```{r load_data}
#| code-fold: true
load("data/raw_nfl.Rdata")
```

We start with the raw 4th down data from the nflverse package. The first six rows and eight columns can be seen here.

```{r head}
#| code-fold: true
head(raw_data)
```

There are a large number of columns in this data, taken from the 2022 NFL season, we do not need all 383. The purpose of this data is to help determine how coaches today approach 4th down, a basic yet essential decision in the NFL. To do this we only need to look at these columns:

```{r filter}
#| code-fold: true
fourth_downs <- raw_data %>%
    filter(!is.na(go_boost) & !is.na(go)) %>%
    select(season, home_coach, away_coach, posteam, defteam, posteam_type, qtr, time, ydstogo, yardline_100, posteam_score, defteam_score, posteam, go_boost, go, epa, wp_fail, wp_succeed, wp, fg_make_prob, miss_fg_wp, make_fg_wp, punt_wp)

kable(head(fourth_downs))
```

Each unit in this table is one 4th down decision. The variables include:

```{r codebook}
#| code-fold: true
codebook <- data.frame("Variable" = names(fourth_downs),
                       "Description" = c("Season",
                                     "Home Team Coach",
                                     "Away Team Coach",
                                     "Offense Team",
                                     "Defense Team",
                                     "Offense Home or Away",
                                     "Quarter",
                                     "Time left in Quarter",
                                     "Yards to Go",
                                     "Yardline Relative to End Zone",
                                     "Offense Score",
                                     "Defense Score",
                                     "Change in Win Probability if Conversion Attempted",
                                     "Conversion Attempted (T/F)",
                                     "Estimated Points Added as a Result of the Play",
                                     "Win Probability if Conversion Fails",
                                     "Win Probability if Conversion Succeeds",
                                     "Win Probability at Time of Play",
                                     "Chance of Making Field Goal",
                                     "Win Probability if Field Goal Misses",
                                     "Win Probability if Field Goal is Made",
                                     "Win Probability if Team Punts"))

kable(codebook)

```

# Decisions

Making a table showing when coaches make the correct decisions on 4th down is essential in evaluating their decision making process. This is made easy by the inclusion of the `go_boost` variable, which tells us the change in win probability if the team decides to go for it.

```{r correct_decisions, message = FALSE}
#| code-fold: true
fourth_decisions <- fourth_downs %>%
    mutate(should_go = ifelse((go_boost > 0),
                             (if_else(go == 100, 1, 0)),
                             NA),
          shouldnt_go = ifelse((go_boost < 0),
                             (if_else(go == 0, 1, 0)),
                             NA),
           coach = if_else(posteam_type == "home", home_coach, away_coach),) %>%
    select(season, coach, posteam, ydstogo, yardline_100, go_boost, should_go, shouldnt_go, go, epa) %>%
    group_by(coach, posteam, season) %>%
    summarize(should_go = 
                sum(should_go == 1, na.rm = TRUE) / 
                (sum(should_go == 0 | should_go == 1, na.rm = TRUE)),
              shouldnt_go = 
                sum(shouldnt_go == 1, na.rm = TRUE) / 
                (sum(shouldnt_go == 0 | shouldnt_go == 1, na.rm = TRUE)),
              EPA = mean(epa),
              count = n()) %>%
    filter(count > 50) %>%
  ungroup()

kable(head(fourth_decisions))
```

This data shows when coaches make the correct choice on 4th down, their average estimated points added on 4th downs throughout the season, and the number of 4th down decisions they had to make throughout the season. It shows the percent of times each coach makes the correct decision in two scenarios: when going for it would have a positive effect on their win probability and when going for it would have a negative effect on their win probability.

# By Situation

One of, if not the most important factor in making a decision on fourth down is the position your team is on the field. Looking at this more closely requires a table that shows coaches decisions based on where they are on the field and how many downs they need to gain to get a first down.

```{r by_position, message = FALSE}
#| code-fold: true
fourth_position <- fourth_downs %>%
  mutate(expected_go = ifelse(go_boost > 0, 1, 0)) %>%
  select(yardline_100,
         ydstogo,
         go,
         expected_go,
         go_boost) %>%
  group_by(yardline_100, ydstogo) %>%
  summarize(actual_go = mean(go)/100,
            expected_go = mean(expected_go, na.rm = TRUE),
            count = n(),
            go_boost = mean(go_boost),
            increase_odds = ifelse(go_boost > 0, 1, 0))

kable(head(fourth_position))
```

This table shows us the percentage of times a coach chose to go for it in the given situation. The first row, for example, shows that 80% of the time a coach decided to go for it when they were at 4th and 1 at the 1 yard line - they were 1 yard away from scoring. It also shows is the average increase in win probability if a coach were to go for it. For example, at 4th and 1 at the 1 yard line, going for it increases your win probability by an average of 5%, an incredibly large number.

```{r save_data}
#| code-fold: true
save(fourth_decisions, fourth_downs, fourth_position, file = "data/clean_data.Rdata")
```


# Decision Trees

For our decision tree building, we would like to format our play by play data in a slightly different manner. Mainly, we want to only have the coach of the team on offense as we are evaluating their deicion-making, not the defensive team's coach. Also, we want to drop unnecessary columns and make a singular `score_diff` column, rather than have separate columns for `posteam_score` and `defteam_score`. The resultant table can be seen below, and is used in our decision tree analysis later on in the project. 

```{r}
#| code-fold: true
load("data/raw_nfl.Rdata")
df <- raw_data %>%
    filter(!is.na(go_boost) & !is.na(go)) %>%
    select(season, home_coach, away_coach, posteam, defteam, posteam_type, game_half, half_seconds_remaining, ydstogo, yardline_100, posteam_score, defteam_score, posteam, go_boost, go, epa, wp_fail, wp_succeed, wp, fg_make_prob, miss_fg_wp, make_fg_wp, punt_wp)
df <- df %>%
  mutate(coach = if_else(posteam_type == "home", home_coach, away_coach),
         home_coach = coach,
         score_diff = posteam_score - defteam_score,
         go = if_else(go == 100, 1, 0),
         game_half = if_else(game_half == "Half1", 1, 2)) %>%
  select(-coach,
         -away_coach,
         -season,
         -posteam,
         -defteam,
         -posteam_type,
         -epa,
         -punt_wp,
         -posteam_score,
         -defteam_score) %>%
  rename(coach = home_coach) %>%
  select(-go, everything())
write.csv(df, "data/dt.csv")

kable(head(df))
```