---
title: "Decision Trees Cleaning"
bibliography: reference.bib
editor_options: 
  chunk_output_type: inline
format:
  html:
    toc: false
---

```{r setup, message = FALSE}
#| code-fold: true
#| code-summary: Inputs
library(tidyverse)
library(kableExtra)
```

# Decision Trees

For our decision tree building, we would like to format our play by play data in a slightly different manner. Mainly, we want to only have the coach of the team on offense as we are evaluating their deicion-making, not the defensive team's coach. Also, we want to drop unnecessary columns and make a singular `score_diff` column, rather than have separate columns for `posteam_score` and `defteam_score`. The resultant table can be seen below, and is used in our decision tree analysis later on in the project. 

```{r}
#| code-fold: true
load("data/raw_nfl.Rdata")
df <- raw_data %>%
    filter(!is.na(go_boost) & !is.na(go)) %>%
    select(season, home_coach, away_coach, posteam, defteam, posteam_type, game_half, half_seconds_remaining, ydstogo, yardline_100, posteam_score, defteam_score, posteam, go_boost, go, epa, wp_fail, wp_succeed, wp, fg_make_prob, miss_fg_wp, make_fg_wp, punt_wp)
df <- df %>%
  mutate(coach = if_else(posteam_type == "home", home_coach, away_coach),
         home_coach = coach,
         score_diff = posteam_score - defteam_score,
         go = if_else(go == 100, 1, 0),
         game_half = if_else(game_half == "Half1", 1, 2)) %>%
  select(-coach,
         -away_coach,
         -season,
         -posteam,
         -defteam,
         -posteam_type,
         -epa,
         -punt_wp,
         -posteam_score,
         -defteam_score) %>%
  rename(coach = home_coach) %>%
  select(-go, everything())
write.csv(df, "data/dt.csv")

kable(head(df))
```

