---
title: "4th Down Decisions"
editor_options: 
  chunk_output_type: inline
bibliography: reference.bib
---

```{r setup, include = FALSE}
library(tidyverse)
library(ggplot2)
library(grid)
library(reticulate)
library(kableExtra)
library(gridExtra)
library(plotly)
library(RColorBrewer)
library(scales)
display.brewer.all(type = 'div')
use_condaenv("r-env")
options(scipen = 999)
```

```{r load_data, include = FALSE}
load("clean_data.Rdata")
```

We will initially explore the data in a similar manner to the method seen in David Romer's paper [@romer2006]. In his evaluation of fourth down decisions, he followed a four step process. First, he estimated the value of kicking (not going for it). Second, he estimated the value of going for it. Third, he figured out which decision was better. Finally, he compared this to the actual decisions made by coaches.

Thanks again to the nflverse package, we already have completed steps one and two as we can simply use the `go_boost` variable to estimate the value of kicking or going for it on each fourth down. Step three was already completed in the cleaning process, as we already have a table that takes into account the fact that if `go_boost` is negative, then the offensive team should kick. If `go_boost` is positive, then they should go for it. We have all the information to evaluate theoretical decisions, now we must evaluate the actual decisions made by coaches.

# 4th Down Statistics

How often do coaches make the correct decision on 4th down?

```{r coach_decisions, warning = FALSE, echo = FALSE}
p <- ggplot(data = fourth_decisions, aes(x = should_go)) + 
  geom_histogram(binwidth = 0.05, fill = "#69b3a2", color="#e9ecef") +
  scale_x_continuous(limits = c(0, 1.1), breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(limits = c(0, 70)) +
  ggtitle("When NFL Coaches Go When They Should") +
  xlab("") +
  stat_bin(binwidth = 0.05, 
           geom='text', 
           aes(label=ifelse(..count.. != 0, ..count.., "")), 
           vjust = -0.5)
p1 <- ggplot(data = fourth_decisions, aes(x = shouldnt_go)) + 
  geom_histogram(binwidth = 0.05, fill = "#69b3a2", color="#e9ecef") +
  scale_x_continuous(limits =c(0, 1.1), breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(limits = c(0, 200)) +
  ggtitle("When NFL Coaches Don't Go When They Shouldn't") +
  xlab("") +
  stat_bin(binwidth = 0.05, 
           geom='text', 
           aes(label=ifelse(..count.. != 0, ..count.., "")), 
           vjust = -0.5)

grid.arrange(p, p1, ncol = 1)
```

We can see a clear difference between these two plots when they are put on the same scale. When it would be a "correct" decision to kick rather than go for it (when win probability would go down if a conversion was attempted) coaches often make the correct decision. Every coach was correct in these situations at least 90% which makes sense because these decisions are often easy to make. They are not the borderline decisions that we can learn a lot about coaches from. For those we turn to the first graph, where we can clearly see that the decision on whether or not to go for it when doing so would increase your win probability is much more difficult, as only two coaches got this one right over 50% of the time. Let's look at this decision more closely and see what turns coaches away from going for it even when they should be.

```{r position_decisions_actual, echo = FALSE}
fourth_position <- fourth_position %>%
  filter(count > 10)
plot_ly(data = fourth_position, 
        x = ~yardline_100, 
        y = ~ydstogo,
        text = ~paste("go:", actual_go, "<br>count:", count),
        type = "scatter",
        mode = "markers",
        marker = list(size = 8),
        color = ~actual_go,
        colors = (c("#FF0000", "#ecf54e", "#008000"))) %>%
  layout(title = "When Coaches Actually Go For It",
         xaxis = list(title = "Yardline (Yards from Endzone)",
                      dtick = 10,
                      tick0 = 0,
                      tickmode = "linear"),
         yaxis = list(title = "Yards to Go",
                      dtick = 5,
                      tick0 = 0,
                      tickmode = "linear")) %>%
  colorbar(title = "")
```

------------------------------------------------------------------------

```{r position_decisions_correct, echo = FALSE}
plot_ly(data = fourth_position, 
        x = ~yardline_100, 
        y = ~ydstogo,
        text = ~paste("go:", expected_go, "<br>count:", count),
        type = "scatter",
        mode = "markers",
        marker = list(size = 8),
        color = ~expected_go,
        colors = (c("#FF0000", "#ecf54e", "#008000"))) %>%
  layout(title = "When Coaches Should Go For It",
         xaxis = list(title = "Yardline (Yards from Endzone)",
                      dtick = 10,
                      tick0 = 0,
                      tickmode = "linear"),
         yaxis = list(title = "Yards to Go",
                      dtick = 5,
                      tick0 = 0,
                      tickmode = "linear")) %>%
  colorbar(title = "")

```

------------------------------------------------------------------------

The first graph is colored based on the proportion of plays in which a coach decides to go for it when doing so would have a positive impact on their win probability. The second graph is colored based on the proportion of plays in which going for it would have a positive impact on their win probability, regardless of whether or not a coach actually decided to go for it.

These graphs clearly show that coaches in the NFL are far too conservative. The only situation in which they consistently go for it is on 4th and 1 inside the 50. The data suggest that they should actually always go for it when there are less than 2 yards to go and almost always go for it within 5 yards. Coaches are consistently far too conservative in situations when it is clearly advantageous to attempt a 4th down conversion. Why might this be the case? Let's perform some EDA to come up with some possible explanations.
